\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

  \journal{An awesome journal} % Sets Journal name


\usepackage{lineno} % add
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={A comparison of model-based and design-based approaches for spatial data.},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

% Pandoc citation processing

% Pandoc header

\usepackage{bm} \usepackage{bbm} \usepackage{color} \DeclareMathOperator{\var}{{var}} \DeclareMathOperator{\cov}{{cov}}

\begin{document}
\begin{frontmatter}

  \title{A comparison of model-based and design-based approaches for spatial
data.}
    \author[USEPA]{Michael Dumelle\corref{1}}
   \ead{Dumelle.Michael@epa.gov} 
    \author[STLAW]{Matthew Higham\corref{1}}
   \ead{mhigham@stlaw.edu} 
    \author[OSU]{Lisa Madsen}
  
    \author[USEPA]{Anthony R. Olsen}
  
    \author[NOAA]{Jay M. Ver Hoef}
  
      \address[USEPA]{United States Environmental Protection Agency, 200 SW 35th St,
Corvallis, Oregon, 97333}
    \address[STLAW]{Saint Lawrence University Department of Math, Computer Science, and
Statistics, 23 Romoda Drive, Canton, New York, 13617}
    \address[OSU]{Oregon State University Department of Statistics, 239 Weniger Hall,
Corvallis, Oregon, 97331}
    \address[NOAA]{Marine Mammal Laboratory, Alaska Fisheries Science Center, National
Oceanic and Atmospheric Administration, Seattle, Washington, 98115}
      \cortext[1]{Corresponding Author}
  
  \begin{abstract}
  This is the abstract.
  
  It consists of two paragraphs.
  \end{abstract}
  
 \end{frontmatter}

\emph{Text based on elsarticle sample manuscript, see
\url{http://www.elsevier.com/author-schemas/latex-instructions\#elsarticle}}

\hypertarget{sec:introduction}{%
\section{Introduction}\label{sec:introduction}}

Please leave comments in your
\textcolor{blue}{c}\textcolor{magenta}{o}\textcolor{green}{l}\textcolor{red}{o}\textcolor{cyan}{r}:
\textcolor{blue}{Michael}, \textcolor{magenta}{Matt},
\textcolor{green}{Lisa}, \textcolor{red}{Tony}, \textcolor{cyan}{Jay}.

There are two approaches to statistical inference for spatial data:
design-based and model-based. In the design-based approach, statistical
inference is based on random sampling. In the model-based approach,
statistical inference is based on a set of distributional assumptions
for the data.

In either approach, statistical inference requires estimating parameters
of interest. Often, these parameters are means or totals. In the
design-based approach, parameters correspond to some population. In the
model-based approach, parameters correspond to some statistical process,
sometimes called a superpopulation.

Spatial data paragraph

define sampling design, define parameters, define model-based, sample,
population, super-population?, etc.

introduction of spatial data / definition of spatial and maybe general
background?

give some examples of possible spatial data

2-3 sentences about each design

goals and outline of paper

De Gruijter and Ter Braak (1990) give an early comparison of the
paradigms for spatial data, quashing the belief that classical
sampling-based methods could not be used for spatially correlated data.
Thereafter, several comparisons between model-based and design-based for
spatial data have been considered, but they tend to compare spatial
models to IRS (Independent Random Sampling) or SIRS (stratified IRS)
(Brus and De Gruijter, 1997). No comparison has been made between
spatial models and spatially balanced design-based approaches.

brus2020statistical: most similar to what this paper will be, directly
comparing the two approaches. This has a lot of other potentially useful
references to go back to. (published in 2021 so likely doesn't have many
that have cited it yet).

\begin{itemize}
\item
  hansen1983evaluation - general overview (in non spatial context), talk
  about population vs superpopulation give a set of guiding principles
  in the discussion
\item
  verhoef2002sampling - strictly defines classical sampling, compares
  FPBK to classical, BK to classical
\item
  wang2012spatial - sampling and inference steps, thorough overview in
  terms of spatial context, lots of references
\end{itemize}

In Section \ref{sec:background}

\hypertarget{sec:background}{%
\section{Background}\label{sec:background}}

\hypertarget{design-based-philosphy}{%
\subsection{Design-Based Philosphy}\label{design-based-philosphy}}

Design-Based Overview

The design-based approach to inference uses characteristics of the
sampling design to estimate parameters of interest. The design-based
approach assumes the data, \(\mathbf{Z}\), are fixed. Randomness is
incorporated through the random selection of sampling units based on
probabilities derived using the sampling design itself. Some examples of
commonly used sampling designs include independent random sampling
(IRS), stratified random sampling, and cluster sampling. Viewing the
data as fixed and incorporating randomness through the sampling design
yields estimators having very few assumptions. Asymptotic properties of
these types of estimators are derived using limiting arguments. Means
and totals, for example, are asymptotically Gaussian. Särndal et al.
(2003) and Lohr (2009) provide thorough reviews of the design-based
approach.

\hypertarget{model-based-philosphy}{%
\subsection{Model-Based Philosphy}\label{model-based-philosphy}}

Model-Based Overview

On the other hand, model-based inference imposes additional assumptions
on the data with a potential to provide more precise estimators if the
additional assumptions hold. Instead of estimating true but unknown
parameters, the goal of model-based inference in the spatial context is
often \emph{prediction} of an unknown quantity. This is a fundamental
philosophical difference between sampling-based and model-based
approaches. Instead of \emph{estimating} a fixed unknown mean, we are
\emph{predicting} the value of the mean, a random variable. We know that
if we sampled all sites, we would have an exact prediction for the mean
of our one realized spatial surface, without any uncertainty. But, the
true mean of the spatial process that generated our realized data is
still not known, and, in the prediction context, we typically do not
care much about what value the mean of the underlying process takes.

We only use FPBK in this paper in order to focus more on comparing the
design-based and model-based approaches. However, k-nearest-neighbors
(Fix and Hodges, 1951; Ver Hoef and Temesgen, 2013), random forest
(Breiman, 2001), Bayesian models (Chan-Golston et al., 2020), among
others, can also be used to obtain predictions for a mean or total from
spatially correlated responses in a finite population setting.

Figure 1a. Data is fixed. In a finite population example, show a 3d
surface that can be generated by anything. If we repeatedly sample the
surface, then 95\% of all 95\% CIs will contain the true mean, which
never changes.

Figure 1b. Spatial process is fixed. In a finite population example,
show 10 3d surfaces that are generated from some model. If we repeatedly
generate the surface and obtain a sample, then 95\% of all 95\% PIs will
contain the realized means. The realized mean changes from surface to
surface and it's not necessarily the case that 95\% of all 95\% PIs will
contain the true, underlying mean.

\hypertarget{comparing-design-based-vs.-model-based}{%
\subsection{Comparing Design-Based
vs.~Model-Based}\label{comparing-design-based-vs.-model-based}}

Several comparisons between the two paradigms have been made in many
different contexts. Sterba (2009) give some history of the two paradigms
as well as some applications in psychological contexts. Cooper (2006)
review the two approaches in an ecological context before introducing a
``model-assisted'' variance estimator that combines aspects from each
approach. We note that while there has been substantial research and
development into estimators that use both design and model-based
principles (see e.g.~Cicchitelli and Montanari (2012), Chan-Golston et
al. (2020) for a Bayesian approach, CITE some more of these), our goal
is not to expand upon or improve these methods.

\hypertarget{spatially-balanced-design-and-analysis}{%
\subsection{Spatially Balanced Design and
Analysis}\label{spatially-balanced-design-and-analysis}}

Spatially balanced sampling algorithms use spatial information to obtain
samples spread out in space. Spatially balanced samples are useful
because they tend to yield estimators that are more precise than
estimators constructed from a sampling algorithm that is not spatially
balanced (Barabesi and Franceschi, 2011; Benedetti et al., 2017;
Grafström and Lundström, 2013; Robertson et al., 2013; Stevens Jr and
Olsen, 2004; Wang et al., 2013). To quantify spatial balance, Stevens Jr
and Olsen (2004) proposed statistics based on Voroni polygons. Many
spatially balanced sampling algorithms exist, including the Generalized
Random Tessellation Stratified (Stevens Jr and Olsen, 2004), the Local
Pivotal Method (Grafström et al., 2012; Grafström and Matei, 2018),
Spatially Correlated Poisson Sampling (Grafström, 2012), Balanced
Acceptance Sampling (Robertson et al., 2013), Within-Sample-Distance
(Benedetti and Piersimoni, 2017), and Halton Iterative Partitioning
(Robertson et al., 2018) algorithms. Here we focus on the Generalized
Random Tessellation Stratified (GRTS) algorithm, which has several
attractive properties that we discuss next.

The GRTS algorithm is used to sample from finite and infinite sample
frames. A finite sample frame contains a finite number of sampling units
and is related to a point geometry. An infinite sample frame contains an
infinite number of sampling units and is related to linear and polygon
geometries. Examples of point, linear, and polygon resources include
lake centroids, stream networks, and wetland areas, respectively. In
addition to its applicability for finite and infinite sample frames, the
GRTS algorithm naturally accommodates stratified designs and designs
with unequal selection probabilities. The algorithm has also been used
to select replacement sites using reverse hierarchical ordering (Stevens
Jr and Olsen, 2004). Replacement sites are used to replace sites in the
original sample that cannot be sampled, often as a result of physical
difficulty in reaching the site or landowner denial of access to the
sites. More recently, the GRTS algorithm also accommodates legacy
(historical) sites, minimum distance between sites, and nearest neighbor
replacement sites. The GRTS algorithm is implemented in the
\textbf{\textsf{R}} package \texttt{spsurvey} (Dumelle et al., 2021).

The GRTS algorithm works by mapping two-dimensional space into
one-dimensional space while incorporating some elements of randomness.
First a square bounding box is placed over the sample frame and divided
into four equally sized squares randomly given an level-one address of
0, 1, 2, or 3. Then the inclusion probabilities of the sampling units in
each square are summed. If this sum is at least one in any of the cells,
a set of four sub-squares are placed within each square. Each set of
four sub-squares are randomly given a second address (a level-two
address) of 0, 1, 2, 3. Then the inclusion probabilities in each
sub-square are summed, and if any of these sums are at least one, a
third level of squares are added in the same manner. This hierarchical
addressing continues until the inclusion probabilities in the smallest
squares are all less than 1. The squares are then mapped to the
one-dimensional line in order of their addresses. The length of this
line equals \(n\), the desired sample size (the sum of the inclusion
probabilities). A uniform random variable, denoted by \(u_1\) is
simulated in \([0, 1]\) and placed on this line. The location of \(u_1\)
on the line can be linked to a sampling unit, denoted by \(s_1\). Then
\(u_2 \equiv u_1 + 1\), which can be linked to another sampling unit,
denoted by \(s_2\). Because the GRTS algorithm requires the inclusion
probabilities of the smallest sub-cells to be less than one, \(s_1\) and
\(s_2\) must be distinct. This process continues until the \(n\)
sampling units have been selected as part of the sample. Further details
are provided by (Stevens Jr and Olsen, 2004) and (Dumelle et al., 2021).

The Horvitz-Thompson estimator (Horvitz and Thompson, 1952) and its
continuous analog (Cordy, 1993) yield unbiased estimates of population
totals (and means). If \(\tau\) is a population total, then the
Horvitz-Thompson estimator of \(\tau\), denoted by \(\hat{\tau}_{HT}\),
is given by \begin{align}\label{eq:htest}
  \hat{\tau}_{HT} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align} where \(Z_i\) is the observed value of the \(i\)th sampling
unit, and \(\pi_i\) is the inclusion probability of the \(i\)th sampling
unit. Estimating the variance of \(\hat{\tau}_{HT}\) directly relies on
knowing the probability that \(s_i\) and \(s_j\) are both included in
the sample (second-order inclusion probability), which can be unfeasible
to compute. Furthermore, this variance estimator does not incorporate
spatial locations. To address these challenges, (Stevens Jr and Olsen,
2003) proposed an alternative variance estimator: the local neighborhood
variance estimator. The local neighborhood variance estimator does not
require second-order inclusion probabilities. It also incorporates
spatial locations, which tends to reduce the estimated variance of
compared \(\hat{\tau}_{HT}\) to a variance estimator ignoring spatial
location (Stevens Jr and Olsen, 2003).

\hypertarget{finite-population-block-kriging}{%
\subsection{Finite Population Block
Kriging}\label{finite-population-block-kriging}}

Finite Population Block Kriging (FPBK) is an alternative to
samipling-based methods (Ver Hoef, 2008). FPBK expands the
geostatistical kriging framework to the finite population setting.
Instead of basing inference off of a specific sampling design, we assume
the data were generated by a spatial process with parameters that can be
estimated using the framework of a model.

Ver Hoef (2008) gives details on the theory of FPBK, but some of the
basic principles are summarized below. For a response variable
\(\mathbf{z}\) that can be measured on a finite number of \(N\) sites,
we want to predict some linear function of the response variable,
\(\tau(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}\) is a vector of weights. For example, if we want to
predict the total abundance across all sites, then we would use a vector
of 1's for the weights.

Typically, however, we only have a sample of the \(N\) sites. Denoting
quantities that are part of the sampled sites with a subscript \emph{s}
and quantities that are part of the unsampled sites with a subscript
\emph{u},

\begin{equation}
\begin{pmatrix} \label{equation:Zmarginal}
    \mathbf{z}_s      \\
    \mathbf{z}_u
\end{pmatrix}
=
\begin{pmatrix}
  \mathbf{X}_s    \\
  \mathbf{X}_u
\end{pmatrix}
\bm{\beta} +
\begin{pmatrix}
\bm{\delta}_s    \\
\bm{\delta}_u
\end{pmatrix},
\end{equation} where \(\mathbf{X}_s\) and \(\mathbf{X}_u\) are the
design matrices for the sampled and unsampled sites, respectively, and
\(\bm{\delta}_s\) and \(\bm{\delta}_u\) are random errors for the
sampled and unsampled sites. Denoting
\(\bm{\delta} \equiv [\bm{\delta}_s \,\, \bm{\delta}_u]'\), we assume
that \(E(\bm{\delta})\) = \(\mathbf{0}\).

We also typically assume that there is spatial correlation in
\(\bm{\delta}\), which can be modeled using a covariance function. Many
common choices for this function assume that spatial covariance
decreases with increasing Euclidean distance between sites. The primary
function used throughout the simulations and applications of this
manuscript is the Exponential covariance function: the \(i,j^{th}\)
entry for \(\mathop{\mathrm{{var}}}(\bm{\delta})\) is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = \theta_1\exp(-3h_{i,j}/\theta_2) + \theta_3\mathbbm{1}\{\mathbf{h}_{i,j} = 0\}, 
\end{align} where \(h_{i,j}\) is the distance between sites \(i\) and
\(j\), and \(\bm{\theta}\) is a vector of spatial covariance parameters
of the partial sill \(\theta_1\), the range \(\theta_2\), and the nugget
\(\theta_3\). However, any spatial covariance function could be used in
the place of the Exponential, including functions that allow for
anisotropy {[}pg. 80 - 93{]}(Chiles and Delfiner, 1999).

With the above model formulation, the Best Linear Unbiased Predictor
(BLUP) for \(\tau(\mathbf{b}'\mathbf{z})\)
\(\tau(\mathbf{b}'\mathbf{z})\) is vague\} and its prediction variance
can be computed. While details of the derivation are in (Ver Hoef,
2008), we note here that the predictor and its variance are both
moment-based. Neither require a particular distribution for
\(\mathbf{z}\).

\hypertarget{numerical-study}{%
\section{Numerical Study}\label{numerical-study}}

cases

\textbf{Base Simulations}

\begin{itemize}
\item
  both good: correctly specified model with high correlation
\item
  break model: non-gaussian errors
\item
  break design: small area estimation
\item
  both good?: misspecified covariance model with high correlation
\item
  break both? non-gaussian areas with smaller sample size
\item
  change n or sampling fraction
\item
  model-based: how should sample be drawn?
\end{itemize}

\hypertarget{software}{%
\subsection{Software}\label{software}}

FPBK can be readily performed in \texttt{R} with the \texttt{sptotal}
package (Higham et al., 2020). We use \texttt{sptotal} for both the
simulation analysis and the application, estimating parameters with
Restricted Maximum Likelihood (REML).

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-barabesi2011sampling}{}%
Barabesi, L., Franceschi, S., 2011. Sampling properties of spatial total
estimators under tessellation stratified designs. Environmetrics 22,
271--278.

\leavevmode\hypertarget{ref-benedetti2017spatially}{}%
Benedetti, R., Piersimoni, F., 2017. A spatially balanced design with
probability function proportional to the within sample distance.
Biometrical Journal 59, 1067--1084.

\leavevmode\hypertarget{ref-benedetti2017spatiallyreview}{}%
Benedetti, R., Piersimoni, F., Postiglione, P., 2017. Spatially balanced
sampling: A review and a reappraisal. International Statistical Review
85, 439--454.

\leavevmode\hypertarget{ref-breiman2001random}{}%
Breiman, L., 2001. Random forests. Machine learning 45, 5--32.

\leavevmode\hypertarget{ref-brus1997random}{}%
Brus, D., De Gruijter, J., 1997. Random sampling or geostatistical
modelling? Choosing between design-based and model-based sampling
strategies for soil (with discussion). Geoderma 80, 1--44.

\leavevmode\hypertarget{ref-chan2020bayesian}{}%
Chan-Golston, A.M., Banerjee, S., Handcock, M.S., 2020. Bayesian
inference for finite populations under spatial process settings.
Environmetrics 31, e2606.

\leavevmode\hypertarget{ref-chiles1999geostatistics}{}%
Chiles, J.-P., Delfiner, P., 1999. Geostatistics: Modeling Spatial
Uncertainty. John Wiley \& Sons, New York.

\leavevmode\hypertarget{ref-cicchitelli2012model}{}%
Cicchitelli, G., Montanari, G.E., 2012. Model-assisted estimation of a
spatial population mean. International Statistical Review 80, 111--126.

\leavevmode\hypertarget{ref-cooper2006sampling}{}%
Cooper, C., 2006. Sampling and variance estimation on continuous
domains. Environmetrics: The official journal of the International
Environmetrics Society 17, 539--553.

\leavevmode\hypertarget{ref-cordy1993extension}{}%
Cordy, C.B., 1993. An extension of the horvitz---thompson theorem to
point sampling from a continuous universe. Statistics \& Probability
Letters 18, 353--362.

\leavevmode\hypertarget{ref-de1990model}{}%
De Gruijter, J., Ter Braak, C., 1990. Model-free estimation from spatial
samples: A reappraisal of classical sampling theory. Mathematical
geology 22, 407--415.

\leavevmode\hypertarget{ref-dumelle2021spsurvey}{}%
Dumelle, M., Olsen, A.R., Kincaid, T., Weber, M., 2021. Selecting and
analyzing spatial probability samples in r using spsurvey. Manuscript
Submitted for Publication.

\leavevmode\hypertarget{ref-fix1951discriminatory}{}%
Fix, E., Hodges, J.L., 1951. Discriminatory analysis, nonparametric
discrimination: Consistency properties. USAF School of Aviation
Medicine.

\leavevmode\hypertarget{ref-grafstrom2012spatiallypoisson}{}%
Grafström, A., 2012. Spatially correlated poisson sampling. Journal of
Statistical Planning and Inference 142, 139--147.

\leavevmode\hypertarget{ref-grafstrom2013well}{}%
Grafström, A., Lundström, N.L., 2013. Why well spread probability
samples are balanced. Open Journal of Statistics 3, 36--41.

\leavevmode\hypertarget{ref-grafstrom2012spatially}{}%
Grafström, A., Lundström, N.L., Schelin, L., 2012. Spatially balanced
sampling through the pivotal method. Biometrics 68, 514--520.

\leavevmode\hypertarget{ref-grafstrom2018spatially}{}%
Grafström, A., Matei, A., 2018. Spatially balanced sampling of
continuous populations. Scandinavian Journal of Statistics 45, 792--805.

\leavevmode\hypertarget{ref-higham2020sptotal}{}%
Higham, M., Ver Hoef, J., Bryce, F., 2020. Sptotal: Predicting totals
and weighted sums from spatial data.

\leavevmode\hypertarget{ref-horvitz1952generalization}{}%
Horvitz, D.G., Thompson, D.J., 1952. A generalization of sampling
without replacement from a finite universe. Journal of the American
statistical Association 47, 663--685.

\leavevmode\hypertarget{ref-lohr2009sampling}{}%
Lohr, S.L., 2009. Sampling: Design and analysis. Nelson Education.

\leavevmode\hypertarget{ref-robertson2013bas}{}%
Robertson, B., Brown, J., McDonald, T., Jaksons, P., 2013. BAS: Balanced
acceptance sampling of natural resources. Biometrics 69, 776--784.

\leavevmode\hypertarget{ref-robertson2018halton}{}%
Robertson, B., McDonald, T., Price, C., Brown, J., 2018. Halton
iterative partitioning: Spatially balanced sampling via partitioning.
Environmental and Ecological Statistics 25, 305--323.

\leavevmode\hypertarget{ref-sarndal2003model}{}%
Särndal, C.-E., Swensson, B., Wretman, J., 2003. Model assisted survey
sampling. Springer Science \& Business Media.

\leavevmode\hypertarget{ref-sterba2009alternative}{}%
Sterba, S.K., 2009. Alternative model-based and design-based frameworks
for inference from samples to populations: From polarization to
integration. Multivariate behavioral research 44, 711--740.

\leavevmode\hypertarget{ref-stevens2003variance}{}%
Stevens Jr, D.L., Olsen, A.R., 2003. Variance estimation for spatially
balanced samples of environmental resources. Environmetrics 14,
593--610.

\leavevmode\hypertarget{ref-stevens2004spatially}{}%
Stevens Jr, D.L., Olsen, A.R., 2004. Spatially balanced sampling of
natural resources. Journal of the american Statistical association 99,
262--278.

\leavevmode\hypertarget{ref-verhoef2008spatial}{}%
Ver Hoef, J.M., 2008. Spatial methods for plot-based sampling of
wildlife populations. Environmental and Ecological Statistics 15, 3--13.

\leavevmode\hypertarget{ref-ver2013comparison}{}%
Ver Hoef, J.M., Temesgen, H., 2013. A comparison of the spatial linear
model to nearest neighbor (k-nn) methods for forestry applications. PloS
one 8, e59129.

\leavevmode\hypertarget{ref-wang2013design}{}%
Wang, J.-F., Jiang, C.-S., Hu, M.-G., Cao, Z.-D., Guo, Y.-S., Li, L.-F.,
Liu, T.-J., Meng, B., 2013. Design-based spatial sampling: Theory and
implementation. Environmental modelling \& software 40, 280--288.


\end{document}


