---
title: "simulation-vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{simulation-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(DvMsp)
```

## Specifying Parameters for a Simulation

We discuss simulations from 36 combinations of parameters in __NAME of paper__, but practitioners may be interested in other combinations that more closely reflect their population of interest. The purpose of this vignette is to illustrate how to use the functions in the `DvMsp` package to investigate performance of model-based predictors and design-based estimators in IRS and GRTS samples for any parameter combination. 

There are two primary functions: `sim_pop()` to simulate a population and `sim_trial()` to perform IRS and GRTS samples and design-based and model-based analyses on those samples. 

The parameters that can be changed in `sim_pop()` are:

* `N`, the number of units in the population,
* `gridded`, whether the population units are placed in a grid (`TRUE`) or are randomly placed (`FALSE`),
* `cortype` the correlation function, which is `"Exponential"` by default.
* `psill`, the partial sill,
* `erange`, the effective range,
* `nugget` the nugget,
* `resptype`, either `"normal"` for Gaussian errors or `"lognormal"` for exponentiated normal errors.


