---
title: A comparison of design-based and model-based approaches for finite population spatial data -- Supporting information.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Jay M. Ver Hoef
    affiliation: NOAA
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Lisa Madsen
    affiliation: OSU
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: St. Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
footnote:
  - code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
journal: "Methods in Ecology and Evolution"
date: "`r Sys.Date()`"
linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \usepackage{bbm}
  \usepackage{color}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
  \renewenvironment{abstract}{}{}
output: rticles::elsevier_article
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  warning = FALSE, 
  message = FALSE,
  include = TRUE, 
  echo = FALSE
)
library(tidyverse)
library(here)
library(xtable) # latex tables
```

Recall the four sampling-analysis combinations from the simulated and real data: simple random sampling with design-based inference (SRS-DB), simple random sapling with model-based inference (SRS-MB), GRTS sampling with design-based inference (GRTS-DB), and GRTS sampling with model-based inference (GRTS-MB). 

# Simulated Data {#sec:simtabs}

For the simulated data, we considered 36 parameter configurations -- the crossing of three sample sizes ($n = 50$, $n = 100$, $n = 200)$, two location layouts (random and gridded), three proportions of dependent random error (DRE) (0%, 50%, 90%), and two response types (normal and skewed). 

Next we present tables summarizing mean bias, RMSE, and interval coverage for all 36 simulation scenarios.
```{r}
make_sim_table <- function(table_value) {
  # browser()
  sim_table <- read_csv(here("inst", "output", "simulation_summary", paste0("allsims", table_value, ".csv")))
  sim_table <- sim_table %>%
    mutate(psill_ratio = paste0(100 * psill_ratio, "%")) %>%
    dplyr::select(Approach = approach,
                  Layout = gridded,
                  Response = resptype,
                  `DRE%` = psill_ratio,
                  n = n,
                  MB = mpbias,
                  RMSE = rmspe,
                  Coverage = coverage
                  ) %>%
    mutate(
     Approach = case_when(
       Approach == "Design GRTS" ~ "GRTS-DB",
       Approach == "Design IRS" ~ "SRS-DB",
       Approach == "Model GRTS" ~ "GRTS-MB",
       Approach == "Model IRS" ~ "SRS-MB"
     ),
     Layout = case_when(
        Layout ~ "Gridded",
        !Layout ~ "Random"
      ),
     Response = case_when(
       Response == "normal" ~ "Normal",
       Response == "lognormal" ~ "Skewed"
     ),
     across(c(`DRE%`, n), as.character)
    )
sim_table <- sim_table[c(2, 4, 1, 3), ]
sim_table_out <- xtable(
  sim_table,
  digits = 4,
  caption = paste0("Sampling-inference combination (Approach), population layout (Layout), response type (Response), proportion of dependent random error (DRE\\%), sample size (n), mean bias (MB), root-mean-squared error (RMSE), and 95\\% interval coverage (Coverage) in simulation scenario ", table_value, "."),
   type = "latex",
  latex.environments = "center",
  align = c("llllrrrrr")
)
print(sim_table_out, include.rownames = FALSE, comment = FALSE, scalebox = "0.8")
}
```

```{r, results = "asis"}
for (i in 1:36) {
  make_sim_table(i)
}
```

\clearpage

# National Lakes Assessment Data {#sec:datatabs}

Next we present tables summarizing mean bias, RMSE, and interval coverage for all six data scenarios.
```{r}
make_data_table <- function(table_value) {
  # browser()
  data_table <- read_csv(here("inst", "output", "data_summary", paste0("datasims", table_value, ".csv")))
  data_table <- data_table %>%
    dplyr::select(Approach = approach,
                  Response = var,
                  n = n,
                  MB = mpbias,
                  RMSE = rmspe,
                  Coverage = coverage
                  ) %>%
    mutate(
     Approach = case_when(
       Approach == "Design GRTS" ~ "GRTS-DB",
       Approach == "Design IRS" ~ "SRS-DB",
       Approach == "Model GRTS" ~ "GRTS-MB",
       Approach == "Model IRS" ~ "SRS-MB"
     ),
     Response = case_when(
       Response == "MMI_ZOOP_NLA6" ~ "ZMMI",
       Response == "TOTALHG_RESULT" ~ "Hg ppb"
     ),
     across(c(n), as.character)
    )
data_table <- data_table[c(2, 4, 1, 3), ]
data_table_out <- xtable(
  data_table,
  digits = 4,
  caption = paste0("Sampling-inference combination (Approach), response type (Response), sample size (n), mean bias (MB), root-mean-squared error (RMSE), and 95\\% interval coverage (Coverage) in a data application scenario ", table_value, "."),
   type = "latex",
  latex.environments = "center",
  align = c("lllrrrr")
)
print(data_table_out, include.rownames = FALSE, comment = FALSE, scalebox = "0.8")
}
```

```{r, results = "asis"}
for (i in 1:6) {
  make_data_table(i)
}
```
