---
title: A comparison of design-based and model-based approaches for finite population spatial data -- Supporting information.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Lisa Madsen
    affiliation: OSU
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Jay M. Ver Hoef
    affiliation: NOAA
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
footnote:
  - code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
journal: "Methods in Ecology and Evolution"
date: "`r Sys.Date()`"
linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \usepackage{bbm}
  \usepackage{color}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
  \renewenvironment{abstract}{}{}
output: rticles::elsevier_article
---

# Simulation study tables {#sec:simtabs}

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  warning = FALSE, 
  message = FALSE,
  include = TRUE, 
  echo = FALSE
)
library(tidyverse)
library(here)
library(xtable) # latex tables
```

Recall the four sampling-analysis combinations from the simulation study, summarized in Table \ref{tab:designanalysis}. Also recall the 36 parameter configurations from the simulation study, summarized in Table \ref{tab:parmtab}.

```{r, echo = FALSE, results = "asis"}
tab <- matrix(c("IRS-Design", "IRS-Model", "GRTS-Design", "GRTS-Model"),
       nrow = 2, byrow = TRUE) %>%
  as.table()
row.names(tab) <- c("IRS", "GRTS")
xtab1 <- xtable(tab, col.names = c("Design", "Model"), caption = "\\label{tab:designanalysis} Sampling-analysis combinations in the simulation study. The rows give the two types of sampling designs and the columns give the two types of analyses.")
align(xtab1) <- "r|ll"
names(xtab1) <- c("Design", "Model")
print(xtab1, comment = FALSE)
## I really want "Analysis" to appear as a row header and "Sampling Design" 
## to appear as a column header, but this isn't easy to figure out.
## I couldn't figure this out with xtable either so it might need to be
## made "by hand."
```


```{r, echo = FALSE, results = "asis"}
n <- c("50", "100", "200")
site_locations <- c("Random", "Gridded", "")
psill_ratio <- c("0", "0.5", "0.9")
resp_type <- c("Normal", "Lognormal", "")
parm_tab <- rbind(n, site_locations, psill_ratio, resp_type)
parm_tab[c(2, 4), 3] <- "-"
row.names(parm_tab) <- c("Sample Size (n)", "Layout", 
                         "Proportion of Dependent Error", "Response Type")
# library(knitr) kable with htlm has trouble with page breaks -- kable with latex harder for me than xtable
# kable(parm_tab, col.names = c("", "", ""),
#       caption = "\\label{tab:parmtab} Simulation parameters. Total variability for all scenarios was 2 so that the partial sill was 0, 1, or 1.8.")
parm_xtab <- xtable(parm_tab, caption = "\\label{tab:parmtab} Simulation scenario options. All combinations of sample size, location layout, response type, and proportion of dependent random error composed the 36 simulation scenarios. In each simualtion scenario, the total variance was 2.")
align(parm_xtab) <- "r|lll"
print(parm_xtab, include.colnames = FALSE, hline.after = c(0, nrow(parm_xtab)), comment = FALSE)
```

Next we present tables summarizing mean estimation bias (design-based) or mean prediction bias (model-based), root-mean-squared error (design-based) or root-mean-squared-prediction error (model-based), mean standard errors, and 95% confidence interval coverage (design-based) or 95% prediction interval coverage (model-based) for all 36 simulation scenarios.
```{r}
make_sim_table <- function(table_value) {
  sim_table <- read_csv(here("inst", "output", "simulation_summary", paste0("allsims", table_value, ".csv")))
  sim_table <- sim_table %>%
    dplyr::select(Approach = approach,
                  Layout = gridded,
                  Response = resptype,
                  PDE = psill_ratio,
                  n = n,
                  Bias = mpbias,
                  `rMS(P)E` = rmspe,
                  MStdE = mse,
                  Coverage = coverage
                  ) %>%
    mutate(
     Approach = case_when(
       Approach == "Design GRTS" ~ "GRTS-Design",
       Approach == "Design IRS" ~ "IRS-Design",
       Approach == "Model GRTS" ~ "GRTS-Model",
       Approach == "Model IRS" ~ "IRS-Model"
     ),
     Layout = case_when(
        Layout ~ "Gridded",
        !Layout ~ "Random"
      ),
     Response = case_when(
       Response == "normal" ~ "Normal",
       Response == "lognormal" ~ "Lognormal"
     ),
     across(c(PDE, n), as.character)
    )
sim_table <- sim_table[c(2, 4, 1, 3), ]
sim_table_out <- xtable(
  sim_table,
  digits = 4,
  caption = paste0("Sampling-analysis combination (Approach), population layout (Layout), response type (Response), proportion of dependent random error (PDE), sample size (n), mean (prediction) bias (Bias), root-mean-squared-(prediction) error (rMS(P)E), mean standard error (MStdE), and 95\\% interval coverage (Coverage) in simulation scenario ", table_value, "."),
   type = "latex",
  latex.environments = "center",
  align = c("llllrrrrrr")
)
print(sim_table_out, include.rownames = FALSE, comment = FALSE, scalebox = "0.8")
}
```

```{r, results = "asis"}
for (i in 1:36) {
  make_sim_table(i)
}
```
